\documentclass{article}

% Submission version (double-blind, anonymous)
\usepackage{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{array}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% ----------------------------------------------------------------
% Custom macros
% ----------------------------------------------------------------
% NeurIPS checklist answer macros
\newcommand{\answerYes}[1]{\textbf{[Yes]} #1}
\newcommand{\answerNo}[1]{\textbf{[No]} #1}
\newcommand{\answerNA}[1]{\textbf{[N/A]} #1}

\newcommand{\CDI}{\textsc{cdi}}
\newcommand{\CQI}{\textsc{cqi}}
\newcommand{\PhAQ}{\textsc{phaq}}
\newcommand{\ATC}{\textsc{atc\_cqi}}
\newcommand{\CY}{\textsc{cy}}
\newcommand{\CPP}{\textsc{cpp}}
\newcommand{\CIDI}{\textsc{cidi}}
\newcommand{\PISA}{PISA}
\newcommand{\AEC}{\textsc{aec}}
\newcommand{\EN}{\textsc{en}}
\newcommand{\CS}{\textsc{cs}}

% ----------------------------------------------------------------
\title{Epistemic Role Framing Elicits Deep Collaboration in LLM Multi-Agent Systems}

% Anonymous submission — author block omitted
\author{Anonymous Authors}

\begin{document}

\maketitle

% ================================================================
\begin{abstract}
Evaluating whether LLM agents genuinely collaborate---rather than merely
exchange outputs---demands metrics that go beyond task accuracy to
characterise \emph{how} agents reason together. We address this measurement gap with four contributions: (1) the
Collaborative Phase Profile (\CPP) evaluation framework,
grounded in the PISA 2015 CPS model, producing four metrics (\CDI,
\CQI, \PhAQ, \CY); (2) the Collaborative Interdependence-Directed
Information (\CIDI) benchmark---$140$ MATH problems with
\CIDI-generated splits verified by \AEC{} to achieve both-solo-zero in
$85\%$ of problems; (3) a large-scale epistemic role framing
study on the benchmark, conditional on CIDI-amenable problems; and
(4) the Agent Epistemic Contribution (\AEC) Shapley
decomposition distinguishing structural from activated interdependence.
Using this infrastructure, we investigate \emph{epistemic role framing}:
does instructing agents to adopt the communicative stance of active
learners---rather than expert solvers---shift behavior from
answer-generation toward collaborative sense-making? We compare three
conditions across $1{,}967$ conversations: C2 (natural),
C6 (peer-aware), and C7 (student role).
Student-role framing drives a decisive shift: mean \CDI{} rises from
$0.362$ to $0.613$ ($\Delta = +0.251$, $d = 1.30$, $p < 0.001$) and
fully-coupled collaboration (COUPLING) more than doubles ($11\% \to
24\%$). Peer-awareness alone (C6) yields no practically meaningful
improvement ($d = 0.16$, not significant after Bonferroni correction),
suggesting that richer framing content is required to produce measurable
collaborative depth in this regime. C7 numerically
outperforms C2 across all 30 subject $\times$ difficulty cells, though
per-cell sample sizes are small and this is descriptive evidence rather
than formal cell-level generalization.
A Shapley-value decomposition (\AEC) further distinguishes
\emph{structural} interdependence---encoded by the \CIDI{} split
($85\%$ both-solo-zero)---from \emph{activated} epistemic necessity,
achieved in $25\%$ of problems even under the strongest tested framing
(pair correctness $3.6\times$ solo maximum). This structural vs.\
activated distinction provides a principled vocabulary for diagnosing
benchmark difficulty and framing effects. We release the \CPP{} rubric,
\CIDI{} pipeline, conversation logs, \AEC{} scripts, ablation transcripts,
and a benchmark card documenting corpus conditionality and intended use.
\end{abstract}

% ================================================================
\section{Introduction}
\label{sec:intro}

Multi-agent systems built on LLMs have demonstrated remarkable
capabilities in code generation \citep{hong2023metagpt}, scientific
reasoning \citep{lu2023mathvista}, and debate-based
refinement \citep{liang2023encouraging}. In most of these systems,
collaboration is designed at the \emph{task level}: agents receive
different tools, different roles in a pipeline, or access to different
data shards. Relatively little attention has been paid to whether the
\emph{epistemic stance} an agent adopts---how it positions itself
cognitively with respect to its own knowledge and its partner's
knowledge---can alter the quality of emergent collaboration.

This question has both evaluation and design implications. On the
evaluation side, benchmarking collaborative agents requires metrics
that can distinguish genuine joint reasoning from efficient
answer-delegation---a distinction that task-accuracy scores obscure.
On the design side, research in the learning sciences has long
established that productive collaboration requires participants to
engage in genuine co-construction of understanding rather than
answer-sharing or expertise-transfer
\citep{roschelle1995construction, dillenbourg1999collaborative}.
If LLM agents default to the role of a knowledgeable assistant---which
is precisely what instruction fine-tuning optimises for---eliciting
collaborative behaviour requires an explicit epistemic intervention,
not merely a task decomposition. Developing and validating such
interventions demands evaluation infrastructure that does not yet
exist for LLM multi-agent systems.

We operationalise this intervention through three increasingly
specific framing conditions (see Section~\ref{sec:design}). The
weakest framing (C2) provides no special instructions beyond
handing each agent a partial information packet. The intermediate
framing (C6) additionally makes agents \emph{aware} that a
peer with complementary information exists. The strongest framing (C7) instructs agents to adopt the
communicative stance of an active student: express uncertainty
explicitly, ask clarifying questions, verify understanding aloud before
advancing to solution steps, and treat the partner's contributions as
material to reason about.

To quantify how deeply agents collaborate, we introduce the
Collaborative Phase Profile (\CPP) framework
(Section~\ref{sec:framework}), grounded in the PISA 2015 CPS
competency model \citep{pisa2015cps}. \CPP{} decomposes each
conversation into a $4 \times 3$ matrix of collaborative phases and
competencies and derives four summary statistics: \CDI{} (breadth of
phase engagement), \CQI{} (quality-weighted depth), \PhAQ{} (quality
of joint exploration), and \CY{} (composite yield).

\paragraph{Contributions.} This paper makes four contributions to the
evaluation and design of LLM multi-agent systems:
\begin{enumerate}
    \item \textbf{The \CPP{} evaluation framework}: a principled,
    reproducible scoring rubric for measuring collaborative depth in
    LLM multi-agent dialogues, grounded in the PISA 2015 CPS competency
    model. \CPP{} provides four complementary metrics (\CDI, \CQI,
    \PhAQ, \CY) and a quadrant taxonomy (COUPLING / PROD\_FAIL /
    TRIVIAL / COLLAPSE) that jointly characterise both process quality
    and outcome.
    \item \textbf{The \CIDI{} benchmark}: an interdependence-controlled
    corpus of $140$ MATH problems with \CIDI-generated packets designed
    to make collaboration structurally necessary, verified by \AEC{}
    to yield both-solo-zero in $85\%$ of problems. The corpus enables
    controlled comparison of framing conditions on problems where
    information-split interdependence is empirically confirmed.
    \item \textbf{Controlled evidence for epistemic role framing}: a
    large-scale ($n = 1{,}967$ conversations, plus 832 ablation
    conversations) study showing that active-learner framing shifts
    behavior from expert-answering to collaborative sense-making
    ($d = 1.30$, $p < 0.001$, conditional on CIDI-amenable problems),
    while peer-awareness alone does not ($d = 0.16$, not significant
    after Bonferroni correction), and that neither framing nor split alone
    suffices---only their combination produces the full effect.
    \item \textbf{Structural vs.\ activated interdependence via \AEC}:
    a Shapley-value decomposition that distinguishes whether
    interdependence is merely present in the task design (structural:
    $85\%$ both-solo-zero) versus actually exploited under the strongest
    tested framing (activated under C7: $25\%$ epistemic necessity rate;
    collaborative surplus $+0.207$)---a diagnostic applicable to any
    multi-agent benchmark using information splitting.
\end{enumerate}

% ================================================================
\section{Background}
\label{sec:background}

\subsection{PISA 2015 Collaborative Problem Solving Framework}

The OECD's PISA 2015 assessment introduced a widely cited framework
that characterises collaborative problem solving along two axes: the
\emph{problem-solving process} (Exploring \& Understanding~[A],
Representing \& Formulating~[B], Planning \& Executing~[C], and
Monitoring \& Reflecting~[D]) and the \emph{collaboration competency}
(Establishing \& Maintaining Shared Understanding~[1], Taking
Appropriate Action to Solve~[2], and Establishing \& Maintaining Team
Organisation~[3]) \citep{pisa2015cps}. Their cross-product defines 12
cells---e.g., \textit{A1}: jointly building an initial problem model,
\textit{D3}: collectively revising the team's shared strategy. PISA
operationalised these cells for human dyads; we adapt the framework to
LLM conversations with minor modifications described in
Section~\ref{sec:framework}.

\subsection{LLM Multi-Agent Systems}

The design space for LLM multi-agent collaboration has expanded
rapidly. Role-based differentiation (e.g., GPT-4 as planner vs.\
executor \citep{park2023generative}), debate protocols that force
agents to refine each other's outputs \citep{liang2023encouraging},
and jigsaw information splitting \citep{du2023improving} each
represent distinct mechanisms. Our work is closest to jigsaw designs,
but differs in two respects: (1) we generate splits that are
semantically interdependent at the \emph{reasoning} level rather than
merely partitioning facts, and (2) we study the independent effect of
agent framing orthogonal to task structure.

\subsection{Epistemic Interdependence and Communicative Role Framing}

Epistemic interdependence---the condition where no single participant
possesses sufficient information or reasoning capacity to reach the
goal alone---is a well-established predictor of productive
collaboration in human learning \citep{barron2003conditions,
johnson2009cooperative}. However, creating genuine interdependence in
LLM systems is non-trivial: a capable LLM may infer missing
information or bypass the collaborative requirement entirely. We
address this through the \CIDI{} design (Section~\ref{sec:cidi}).

Communicative role framing---whereby adopting a learner rather than an
expert stance activates different discourse strategies---has been studied
in peer tutoring \citep{chi2001learning} and collaborative inquiry
contexts \citep{king1992facilitating}. To our knowledge, no prior work
has systematically tested whether such framing instructions shift
collaborative behaviour in LLM-to-LLM interactions.

% ================================================================
\section{The CPP Framework}
\label{sec:framework}

\subsection{The 12-Cell Collaborative Phase Matrix}

We adapt the PISA CPS $4 \times 3$ grid to LLM-to-LLM mathematical
dialogues. Each cell $(P, C)$ for process $P \in \{A, B, C, D\}$ and
competency $C \in \{1, 2, 3\}$ is annotated by a trained scorer on a
0--3 scale using a detailed rubric (see Appendix~\ref{app:rubric}):
\textbf{0} = absent, \textbf{1} = attempted but superficial,
\textbf{2} = substantive but incomplete, \textbf{3} = full and
well-articulated. A cell is considered \emph{active} if its score
$\geq 1$.


\subsection{Summary Metrics}

\paragraph{Collaborative Depth Index (\CDI).}
\begin{equation}
  \CDI = \frac{|\{(P,C) : \text{score}(P,C) \geq 1\}|}{12} \in [0, 1].
\end{equation}
\CDI{} measures \emph{breadth}: what fraction of the 12 cells were
genuinely engaged. A value of $1.0$ means every phase-competency
combination was visited; $0.0$ means the dialogue was entirely
unilateral.

\paragraph{Collaborative Quality Index (\CQI).}
\begin{equation}
  \CQI = \frac{\sum_{P,C} \text{score}(P,C)}{36} \in [0, 1].
\end{equation}
\CQI{} weights by quality: 36 is the maximum possible total score
($12 \times 3$). Unlike \CDI{}, \CQI{} penalises superficial
engagement.

\paragraph{Phase A Quality (\PhAQ).}
\begin{equation}
  \PhAQ = \frac{\sum_{C \in \{1,2,3\}} \text{score}(A, C)}{9} \in [0, 1].
\end{equation}
Process~A is the locus of joint problem-model construction; $\PhAQ = 0$ indicates agents proceeded to solution without any joint exploration.

\paragraph{ATC\textsubscript{CQI} (\ATC).}
We also score against the five ATC21S collaborative competency dimensions \citep{griffin2012assessment} (0--3 each); aggregate $\ATC = \frac{\sum 5\text{ dims}}{15}$.

\paragraph{Collaborative Yield (\CY).}
\begin{equation}
  \CY = 0.35 \cdot \CDI + 0.45 \cdot \mathbb{1}[\text{correct}]
        + 0.20 \cdot \delta_{\text{couple}},
\end{equation}
where $\delta_{\text{couple}} = 1$ iff the conversation is in the COUPLING quadrant.

\subsection{Quadrant Taxonomy}

We embed conversations in a 2D outcome space defined by \CDI{}
(process) and binary correctness (outcome):

\begin{center}
\begin{tabular}{lcc}
  \toprule
  \textbf{Quadrant} & \textbf{CDI} & \textbf{Correct} \\
  \midrule
  COUPLING  & $\geq 0.5$ & Yes \\
  PROD\_FAIL & $\geq 0.5$ & No  \\
  TRIVIAL    & $< 0.5$   & Yes \\
  COLLAPSE   & $< 0.5$   & No  \\
  \bottomrule
\end{tabular}
\end{center}

COUPLING (high \CDI, correct) is the ideal---deep collaboration \emph{and} success.
PROD\_FAIL (high \CDI, incorrect) is productive failure \citep{kapur2016examining}.
TRIVIAL (low \CDI, correct) implies one agent solved independently.
COLLAPSE (low \CDI, incorrect) is complete failure of both process and outcome.

\subsection{The CIDI Information Split}
\label{sec:cidi}

A core challenge in designing jigsaw multi-agent tasks is ensuring
that the split is genuinely interdependent at the reasoning level,
not merely informational. We introduce Collaborative
Interdependence-Directed Information (\CIDI): an automated pipeline
that uses GPT-4.1 to decompose each problem into three components:

\begin{enumerate}
  \item \emph{Shared context}: the problem statement elements visible
  to both agents, including notation and domain.
  \item \emph{Packet A}: information given exclusively to Agent~A,
  chosen so that Agent~A cannot solve the problem without the
  \emph{reasoning output} (not just the raw facts) of Agent~B.
  \item \emph{Packet B}: the symmetric complement for Agent~B.
\end{enumerate}

The key criterion is \emph{reasoning-output dependency}: neither agent
can progress past initial exploration without the other's deductions.
Unlike negative-instruction (``constitutional'') splits---which LLMs
frequently override using general domain knowledge---\CIDI{} uses
positive, non-inferable complementary constraints.

% ================================================================
\section{Experimental Design}
\label{sec:design}

\subsection{Problem Corpus Selection (Phase 1)}

We drew problems from the MATH benchmark \citep{hendrycks2021math},
spanning six subject areas (Algebra, Geometry, Number Theory,
Prealgebra, Precalculus, and Counting \& Probability) and five
difficulty levels (MATH levels 1--5), yielding a $6 \times 5 = 30$
cell grid. From an initial pool of 300 problems we applied a
\emph{genuine epistemic filter}: we ran all 300 problems under
condition C7 and retained only those for which $\CDI \geq 0.5$ in at
least two of three replications. The purpose of this filter is to
identify problems for which the \CIDI{} split produces genuine
reasoning-level interdependence---problems that are too easy or too
structurally simple tend to be resolved in one agent's opening message
regardless of framing, making them uninformative for studying
collaborative dynamics. The filter retained $140$ problems
($47\%$ retention rate), distributed across all 30 subject-difficulty
cells with a minimum of 1 problem per cell and a mean of 4.7 problems.

\emph{Selection conditionality.}\enspace Because the filter uses C7
performance to define the corpus, \emph{all estimates are conditional on
\CIDI-amenable problems}: absolute CDI values reflect performance on
problems pre-selected for collaborative potential, and within-corpus
effect sizes may overestimate the benefit expected on an unfiltered sample.
Paired comparisons remain internally valid within this regime.
Future work will establish external validity on an independently drawn unfiltered subset.

\subsection{Conditions (Phase 2)}

All conditions use the same model (GPT-4.1), the same \CIDI{} splits,
and the same conversation scaffolding (alternating turn-taking,
maximum 20 turns (10 per agent in a 2-agent alternating exchange),
no access to external tools). The \emph{only}
manipulation is the system-prompt framing given to each agent.

\paragraph{C2: Natural (L1 baseline).}
Each agent receives only its information packet and the shared
context. No framing about partners, learning, or epistemic role.
Agents are aware they must send messages to proceed but receive no
guidance on what kind of collaboration is expected.

\paragraph{C6: Peer-aware (L2).}
Each agent additionally receives a sentence establishing that it is
working with a \emph{partner} who holds complementary information,
and that the goal is to solve the problem together. This activates
awareness of the collaborative situation without specifying a
cognitive stance.

\paragraph{C7: Student-sim (L3).}
Each agent is instructed to adopt the communicative stance of an active
student: express uncertainty explicitly, ask clarifying questions
before proceeding to solution steps, acknowledge when a step is unclear,
and treat the partner's contributions as material to reason about rather
than accept at face value. Agents are told that producing an
authoritative-sounding response is not the goal---shared understanding
is. This framing targets observable discourse behaviour (question-asking,
uncertainty expression, joint model-building) rather than attributing a
belief state to the model.

\paragraph{Scale and replication.}
Each of the 140 problems was run under each of the three conditions,
with between 3 and 7 replications per problem-condition pair using
different random seeds (mean 4.7 replications), for a total of
$\mathbf{1{,}967}$ conversations. Per-problem mean \CDI{} values
(averaged over replications) are used as the unit of analysis for all
inferential tests.

\subsection{Design Scope}
\label{sec:scope}

The present study compares three conditions (C2, C6, C7) chosen to
vary epistemic framing from absent (C2) to peer-aware (C6) to
active-learner (C7). This three-condition design isolates the effect
of partner-awareness but does not disentangle active-learner
role framing from other strong role framings such as expert-solver or
debate framing, nor does it test whether framing effects persist
without the \CIDI{} information split. Two of these ablations---expert-solver
(C\textsubscript{EXP}) and full-context C7 without the jigsaw split
(C\textsubscript{FULL})---were run on the full 140-problem corpus;
results are reported in Section~\ref{sec:ablations}. A debate/critique
condition (C\textsubscript{DEB}) and an ignorance-only condition remain
as future work.

\subsection{Annotation Protocol}
\label{sec:annotation}

Each conversation was scored by two graduate-student annotators
(neither an author of this paper) trained on the \CPP{} rubric over
two calibration sessions using held-out practice conversations. Annotators
were blinded to condition assignment during scoring; condition labels
were masked and conversations were presented in randomised order.
We note that label masking does not preclude inference from conversation
content; annotators may have implicitly detected framing differences
from discourse style.
Disagreements exceeding one scale point were resolved by consensus
discussion with a third annotator.

\paragraph{Inter-rater reliability.}
Reliability was assessed on a stratified random sample of 36 conversations
(3 conditions $\times$ 4 quadrants $\times$ 3 conversations each).
This constitutes preliminary but condition- and quadrant-balanced
coverage. Table~\ref{tab:reliability} reports four metrics aligned
with the conceptual structure of \CPP{}. Krippendorff's $\alpha$ at
the \emph{ordinal cell level} (0--3 scale per cell) is $0.81$ (range
$0.74$--$0.89$ across the 12 cells), indicating substantial agreement
\citep{hayes2007answering}. $\text{ICC}(2,1)$ for the \CQI{}
score---which averages normalised ordinal scores across cells---is
$0.83$. Cohen's $\kappa$ for the binary \CDI{} threshold ($\geq 0.5$
vs.\ $< 0.5$, derived from the fraction of active cells) is $0.76$.
Four-way quadrant agreement yields $\kappa = 0.72$. All metrics exceed
conventional thresholds for substantial agreement. Expanding the
reliability sample to $\geq 100$ conversations to enable per-cell
$\kappa$ estimates is identified as a key next step.

\begin{table}[t]
  \caption{Inter-rater reliability on 36 stratified conversations
  (preliminary). Metrics correspond to distinct levels of the \CPP{}
  scoring hierarchy.}
  \label{tab:reliability}
  \centering
  \begin{tabular}{llll}
    \toprule
    \textbf{Level} & \textbf{Metric} & \textbf{Target} & \textbf{Value} \\
    \midrule
    Cell ordinal (0--3)     & Krippendorff $\alpha$           & $>0.70$ & $0.81$ \\
    \CQI{} (norm.\ intensity) & ICC$(2,1)$                    & $>0.75$ & $0.83$ \\
    \CDI{} binary ($\geq\!0.5$) & Cohen $\kappa$              & $>0.70$ & $0.76$ \\
    Quadrant (4-way)        & Cohen $\kappa$                  & $>0.60$ & $0.72$ \\
    \bottomrule
  \end{tabular}
\end{table}

% ================================================================
\section{Results}
\label{sec:results}

\subsection{Hypothesis 1: Framing Drives Collaborative Depth}

Table~\ref{tab:main_results} reports mean scores for all metrics
across the three conditions. C7 yields dramatically higher \CDI{} than
both C2 ($\Delta = +0.251$) and C6 ($\Delta = +0.222$). The \CQI{}
increase is similarly large ($+0.106$ over C2), as is the improvement
in \ATC{} ($+0.096$). The \CY{} composite rises by $+0.134$ over the
baseline. All pairwise differences between C7 and C2, and between C7
and C6, are statistically significant with Wilcoxon signed-rank tests
on per-problem mean \CDI{} values ($n = 140$ pairs per comparison;
Table~\ref{tab:stats}).

Critically, the C6 vs.\ C2 comparison yields a Cohen's $d$ of only
$0.16$ ($p = 0.026$), which does \emph{not} survive Bonferroni
correction for three comparisons ($\alpha/3 = 0.0167$). The effect is
negligible in practical terms. Peer-awareness alone does not shift
agents away from expert-answering behavior---it only marginally changes
\CDI{} relative to the natural baseline. This rules out simple partner-awareness as a sufficient explanation,
supporting the interpretation that the active-learner communicative stance
is a key operative component (see also Section~\ref{sec:ablations}).

\begin{table}[t]
  \caption{Mean \CPP{} metric scores by condition. C7 outperforms both
  baselines on all metrics. The C6 vs.\ C2 gap is small ($d=0.16$),
  consistent with active-learner communicative stance---not mere
  partner-awareness---as the key framing component in this regime.}
  \label{tab:main_results}
  \centering
  \begin{tabular}{lrrrrr}
    \toprule
    \textbf{Condition} & \textbf{\CDI} & \textbf{\CQI} & \textbf{\PhAQ} & \textbf{\ATC} & \textbf{\CY} \\
    \midrule
    C2 (L1, Natural)   & 0.362 & 0.144 & 0.053 & 0.404 & 0.275 \\
    C6 (L2, Peer-aware) & 0.390 & 0.156 & 0.060 & 0.430 & 0.304 \\
    C7 (L3, Student-sim) & \textbf{0.613} & \textbf{0.250} & \textbf{0.135} & \textbf{0.500} & \textbf{0.409} \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[t]
  \caption{Pairwise hypothesis tests for \CDI{} (primary outcome).
  Wilcoxon signed-rank tests on per-problem mean \CDI{}, paired by
  problem ($n=140$ pairs per comparison). Effect sizes are Cohen's $d$
  on paired differences. $^\dagger$C6 vs.\ C2 is significant at $\alpha=0.05$
  uncorrected but does \emph{not} survive Bonferroni correction
  ($\alpha/3 = 0.0167$; $p = 0.026 > 0.0167$).}
  \label{tab:stats}
  \centering
  \begin{tabular}{llrrr}
    \toprule
    \textbf{Comparison} & \textbf{Direction} & \textbf{Cohen's $d$} & \textbf{$p$-value} & \textbf{Significant} \\
    \midrule
    C7 vs.\ C2 & C7 $>$ C2 & 1.30 & $< 0.001$ & \checkmark \\
    C7 vs.\ C6 & C7 $>$ C6 & 1.24 & $< 0.001$ & \checkmark \\
    C6 vs.\ C2 & C6 $>$ C2 & 0.16 & 0.026       & $^\dagger$ \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Hierarchical Model: Controlling for Difficulty and Subject}
\label{sec:lmm}

To address the nested structure (replications within problems within
subject-difficulty cells), we fit a linear mixed-effects model with
random intercepts per problem and fixed effects for condition,
difficulty level, and subject area. Condition effects are qualitatively
unchanged from the Wilcoxon tests: C7 ($\hat{\beta} = +0.251$,
$p < 0.001$), C6 ($\hat{\beta} = +0.028$, $p = 0.011$, not significant
after Bonferroni correction). The problem random intercept accounts for
$31\%$ of total \CDI{} variance---non-negligible, and controlling for
it strengthens rather than weakens the condition effect estimates.

\subsection{Hypothesis 2: Phase A Activation Is the Proximal Mechanism}

The \PhAQ{} metric isolates the quality of the joint exploration
phase. Under C2, $67.9\%$ of conversations included any Phase~A
activity ($\PhAQ > 0$); under C6 this rises modestly to $74.3\%$.
Under C7, however, $99.3\%$ of conversations exhibit substantive Phase~A
engagement (139 of 140 problems; $d = 1.22$ for C7 vs.\ C2,
$p < 0.001$). Mean \PhAQ{} under C7 ($0.135$) is $2.5\times$ that
under C2 ($0.053$), indicating not just more frequent Phase~A activity
but higher quality joint exploration.

Phase~A activation is the strongest \emph{observable mediator} of the
C7 framing effect: the active-learner stance makes presenting an
immediate solution communicatively inappropriate, requiring joint
negotiation of shared understanding first. Whether this reflects a
causal mechanism or a prompted discourse pattern requires micro-level
trace analysis.

\subsection{Quadrant Distribution}

Figure~\ref{fig:quadrants} and Table~\ref{tab:quadrants} show the
distribution of conversations across the four outcome quadrants. The
most striking shift is from COLLAPSE to PROD\_FAIL: under C2,
$46\%$ of conversations fall in the COLLAPSE quadrant (low process,
incorrect); under C7, this drops to $18\%$. The COUPLING quadrant
(high process, correct) grows from $11\%$ to $24\%$. The PROD\_FAIL
quadrant grows from $25\%$ to $49\%$, indicating that the student-role
framing substantially increases the rate of productive failure.

The large PROD\_FAIL mass under C7 reflects corpus difficulty, not
framing failure: Phase~1 selected problems that challenge even deep
collaboration. The evaluatively decisive comparison is against COLLAPSE
($46\% \to 18\%$, $2.5\times$ reduction under C7), which signals the
framing prevents unproductive engagement rather than merely
redistributing incorrect outcomes.

\begin{table}[t]
  \caption{Quadrant distribution by condition (percentage of
  conversations). C7 more than doubles COUPLING and reduces COLLAPSE
  by more than $2.5\times$ relative to the natural baseline.}
  \label{tab:quadrants}
  \centering
  \begin{tabular}{lrrrr}
    \toprule
    \textbf{Condition} & \textbf{COUPLING} & \textbf{PROD\_FAIL} & \textbf{TRIVIAL} & \textbf{COLLAPSE} \\
    \midrule
    C2 (Natural)   & 11\% & 25\% & 18\% & 46\% \\
    C6 (Peer-aware) & 13\% & 26\% & 22\% & 39\% \\
    C7 (Student-sim) & \textbf{24\%} & \textbf{49\%} &  9\% & \textbf{18\%} \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
  \begin{axis}[
    ybar stacked,
    bar width=1.4cm,
    width=0.62\textwidth,
    height=4.5cm,
    ymin=0, ymax=100,
    ylabel={Conversations (\%)},
    xtick={1,2,3},
    xticklabels={C2 (Natural), C6 (Peer-aware), C7 (Student-sim)},
    xticklabel style={font=\small},
    legend style={at={(0.5,-0.22)}, anchor=north, legend columns=4,
                  font=\small, column sep=4pt},
    enlarge x limits=0.35,
    ymajorgrids=true,
    grid style={dashed, gray!40},
    tick align=outside,
    every axis plot/.append style={draw=black, line width=0.4pt},
  ]
  % COLLAPSE (bottom, gray) — C2:46, C6:39, C7:18
  \addplot[fill=gray!55] coordinates {(1,46) (2,39) (3,18)};
  % TRIVIAL (yellow) — C2:18, C6:22, C7:9
  \addplot[fill=yellow!70!orange!60] coordinates {(1,18) (2,22) (3,9)};
  % PROD_FAIL (orange) — C2:25, C6:26, C7:49
  \addplot[fill=orange!75] coordinates {(1,25) (2,26) (3,49)};
  % COUPLING (green, top) — C2:11, C6:13, C7:24
  \addplot[fill=green!55!black!80] coordinates {(1,11) (2,13) (3,24)};
  \legend{COLLAPSE, TRIVIAL, PROD\_FAIL, COUPLING}
  \end{axis}
  \end{tikzpicture}
  \caption{Quadrant distribution across conditions. Bars stack from
  bottom: COLLAPSE (gray), TRIVIAL (yellow), PROD\_FAIL (orange),
  COUPLING (green). The dominant shift from C2/C6 to C7 is a transfer
  of mass from COLLAPSE to PROD\_FAIL and COUPLING, driven by the
  Phase~A activation effect (Section~\ref{sec:results}).}
  \label{fig:quadrants}
\end{figure}

\subsection{Epistemic Decomposition: Agent Epistemic Contribution}
\label{sec:aec}

The metrics presented so far measure \emph{how much} agents collaborate
(\CDI) and \emph{how deeply} they engage across phases (\CQI, \PhAQ).
They do not answer a more fundamental question: was collaboration
epistemically \emph{necessary}---that is, did the pair achieve a result
neither agent could reach alone? We address this via a Shapley-value
decomposition we term Agent Epistemic Contribution (\AEC).

\paragraph{Setup.}
For each of the $140$ benchmark problems we run each agent in isolation
with only its own \CIDI{} packet, using the same model (GPT-4.1) and
temperature ($T = 0$) for deterministic baselines (collaborative
sessions used $T = 0.7$ with different random seeds), and the same
maximum token budget per turn. Three value functions are defined,
each on a partial-credit scale $\{0, 0.5, 1\}$:
$v(\{A\})$, $v(\{B\})$, and $v(\{A,B\}) = \text{best C7 replication correctness}$.
Using the best C7 replication as $v(\{A,B\})$ makes \EN{} and \CS{} upper-bound
estimates of activated interdependence under C7. Standard Shapley values for $N=2$ give
$\AEC_A = \tfrac{1}{2}v(\{A\}) + \tfrac{1}{2}(v(\{A,B\}) - v(\{B\}))$
(and symmetrically for $\AEC_B$). Three aggregate measures: \EN{}
($v(\{A,B\}) > \max(v(\{A\}), v(\{B\}))$, pair surpasses both solos);
\CS{} ($v(\{A,B\}) - \max(v(\{A\}), v(\{B\}))$, gain magnitude);
EB ($1 - |\AEC_A - \AEC_B|$, contribution symmetry).
The decomposition separates \emph{structural} interdependence (task design)
from \emph{activated} interdependence (actually exploited under framing)---a
distinction invisible to \CDI{} alone.

\paragraph{Results.}
Table~\ref{tab:aec} summarises the decomposition. Solo correctness is
very low: $v_A = 0.100$, $v_B = 0.057$---consistent with the \CIDI{}
design goal that neither packet alone is solvable. The C7 pair achieves
$v_{AB} = 0.357$, a $3.6\times$ improvement over the better individual.
Critically, $85\%$ of problems exhibit \emph{both-solo-zero}: neither
packet provides sufficient information for any correct answer, validating
the \CIDI{} epistemic filter at the correctness level. \EN{} holds in
$25\%$ of problems---cases where collaboration not only was structurally
required but also succeeded. Mean \CS{} $= +0.207$ and mean EB $=
0.857$: when collaboration succeeds, the gains are substantial and
both agents contribute comparably.

\begin{table}[t]
  \caption{\AEC{} decomposition across $140$ problems. \emph{Both-solo-zero}:
  neither agent's packet alone permits a correct answer; \EN: pair
  surpasses the better solo; \CS: collaborative surplus over best solo;
  EB: Shapley balance between agents.}
  \label{tab:aec}
  \centering
  \begin{tabular}{lr}
    \toprule
    \textbf{Metric} & \textbf{Value} \\
    \midrule
    $v_A$: Agent A solo correctness  & 0.100 \\
    $v_B$: Agent B solo correctness  & 0.057 \\
    $v_{AB}$: C7 pair correctness    & 0.357 \\
    \midrule
    Both-solo-zero rate              & 85.0\% \\
    \EN{} rate (epistemic necessity) & 25.0\% \\
    \CS{} mean (collaborative surplus)  & $+0.207$ \\
    EB mean (epistemic balance)      & 0.857  \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Connecting AEC to framing.}
The $60$-percentage-point gap between structural interdependence ($85\%$
both-solo-zero) and activated epistemic necessity ($25\%$ EN under C7)
is invisible to accuracy metrics, and quantifies headroom available to
better framing or coordination mechanisms. High both-solo-zero with low
\EN{} is a diagnostic signal that collaborative task structure is unactivated.

\subsection{Hypothesis 3: Descriptive Robustness Across Subject and Difficulty}

C7 numerically outperforms C2 in all 30 subject $\times$ difficulty
cells (Appendix~\ref{app:heatmap}); per-cell $n$ values (range 1--8,
mean 4.7) are too small for formal cell-level testing, so H3 constitutes
\emph{descriptive robustness, not formal generalization}. Effect sizes
for cells with $n \geq 3$ range from $d = 0.63$ to $d = 2.77$, with
larger effects at higher difficulty levels---consistent with framing
mattering most when coordinated reasoning is genuinely required.

\subsection{Robustness and Validity Checks}
\label{sec:robustness}

Four checks support the validity of the main findings (details in
Appendix~\ref{app:robustness}):
(W2) C6's negligible gain over C2 ($d = 0.16$, $p = 0.026$,
not surviving Bonferroni correction) rules out simple partner-awareness
as a sufficient mechanism driving C7, supporting the specificity of the active-learner
stance; an ignorance-only ablation would further isolate subcomponents.
(W3) \CDI{} and correctness are orthogonal by design
($P(\text{correct} \mid \CDI \geq 0.5) = 0.321$ vs.\ $0.324$;
$\chi^2 = 0.01$, $p = 0.94$); the C7 advantage in COUPLING ($24\%$ vs.\
$11\%$) holds at \emph{equal mean conversation length} (19.9 turns
across all conditions), ruling out a length confound.
(W4) A DAG-weighted \CDI{} variant ($\CDI_w$, weighting
deeper phases higher) preserves identical condition rankings
($\rho > 0.99$ across 1,967 conversations), demonstrating robustness
to the specific cell-weighting choice.
(W5) Bootstrap $95\%$ CI for $\Delta_\CDI$(C7$-$C2)
$= [0.219, 0.284]$, entirely above zero and bounded away from the C6
effect; $n = 140$ provides $>99.9\%$ power at $d = 1.30$.

% ================================================================
\subsection{Ablation Conditions: C\textsubscript{FULL} and C\textsubscript{EXP}}
\label{sec:ablations}

To disentangle the active-learner communicative stance from the \CIDI{}
structural split, we ran two ablation conditions on the full 140-problem
corpus (3 replications each; 832 of 840 planned conversations completed,
8 C\textsubscript{FULL} runs excluded pre-scoring due to content-policy refusals;
imputing them as CDI$=0$ leaves conclusions unchanged):
C\textsubscript{FULL} applies C7 active-learner framing with
full problem context (both agents see the complete problem; no information
split), and C\textsubscript{EXP} applies the \CIDI{} split with
expert-solver framing instead of the active-learner stance.

Table~\ref{tab:ablations} shows mean \CDI{} and quadrant rates across
all five conditions.

\begin{table}[t]
  \caption{Mean \CDI{} and quadrant distribution across main and ablation
  conditions ($n = 140$ problems). Both ablation conditions fall \emph{below}
  C2, the natural-framing baseline, indicating that neither component alone
  accounts for the C7 effect.}
  \label{tab:ablations}
  \centering
  \begin{tabular}{lrrrrr}
    \toprule
    \textbf{Condition} & \textbf{Mean \CDI{}} & \textbf{COUPLING} & \textbf{PROD\_FAIL} & \textbf{TRIVIAL} & \textbf{COLLAPSE} \\
    \midrule
    C7 (CIDI + learner)       & $0.613 \pm 0.014$ & 24\% & 49\% &  9\% & 18\% \\
    C6 (CIDI + peer-aware)    & $0.390 \pm 0.015$ & 13\% & 26\% & 22\% & 39\% \\
    C2 (CIDI + natural)       & $0.362 \pm 0.015$ & 11\% & 25\% & 18\% & 46\% \\
    \midrule
    C\textsubscript{FULL} (learner, no split) & $0.296 \pm 0.018$ & 11\% & 12\% & 49\% & 28\% \\
    C\textsubscript{EXP} (expert + split)     & $0.267 \pm 0.017$ &  5\% & 17\% & 25\% & 53\% \\
    \bottomrule
  \end{tabular}
\end{table}

Both ablation conditions fall \emph{below} C2: C\textsubscript{FULL}
achieves $\CDI = 0.296 \pm 0.018$ ($d = -0.25$ vs.\ C2, paired), with $49\%$
TRIVIAL---agents solve independently when both hold the full problem,
indicating that the active-learner register alone cannot substitute for
genuine information asymmetry. C\textsubscript{EXP} achieves
$\CDI = 0.267 \pm 0.017$ ($d = -0.43$ vs.\ C2), the lowest of all
five conditions, with $53\%$ COLLAPSE---expert-solver framing suppresses
epistemic inquiry even when the \CIDI{} split is in place. The two
ablations are not meaningfully different from each other ($d = 0.12$),
providing evidence that neither component alone accounts for C7's effect
(C\textsubscript{FULL}: $d = 1.31$; C\textsubscript{EXP}: $d = 1.72$
vs.\ C7): the synergy of structural interdependence \emph{and}
active-learner communicative stance is required.

% ================================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Design implications.}
The C6 $\approx$ C2 finding ($d = 0.16$, not significant after Bonferroni correction)
shows that telling agents they have a collaborative partner \citep{liang2023encouraging,
du2023improving} leaves the default expert-answering mode largely intact in the
\CIDI-amenable regime. C\textsubscript{EXP} (expert framing with \CIDI{} split)
produces the highest COLLAPSE rate across all five conditions, supporting the
interpretation that the active-learner stance is the key operative component.
For benchmark designers, the 60-point gap between structural
($85\%$ both-solo-zero) and activated ($25\%$ EN under C7) interdependence
captured by \AEC{} is invisible to accuracy metrics: high both-solo-zero with
low \EN{} signals that framing is failing to exploit the task's collaborative structure.

\paragraph{Ablation results.}
The key synthesis (Section~\ref{sec:ablations}): neither component alone suffices.
C\textsubscript{FULL} falls below C2 ($49\%$ TRIVIAL; identical information removes collaboration pressure).
C\textsubscript{EXP} achieves the lowest \CDI{} of all five conditions ($53\%$ COLLAPSE), showing expert framing suppresses inquiry even with the structural split in place.

\paragraph{Limitations.}
Several limitations bound the current conclusions.
\emph{(L1) Corpus conditionality}: all estimates are conditional on
\CIDI-amenable problems (corpus filtered using C7; see §4.1); effect
sizes may overestimate the benefit expected on an unfiltered sample.
\emph{(L2) Remaining scope}: subcomponent ablations (ignorance-only, debate framing) remain; \AEC{} EN and CS are upper-bound estimates under C7 (best replication), requiring per-condition runs for cross-condition comparison.
\emph{(L4) Model generalisability}: all agents use GPT-4.1; framing
effects may differ for other model families or fine-tuned models.
\emph{(L5) Reliability sample}: the reliability evidence is
preliminary ($n=36$); per-cell $\kappa$ estimates require $\geq 100$
conversations.
\emph{(L6) Collaborative register vs.\ reasoning}: we cannot rule out
that C7 agents produce high-\CDI{} discourse without the underlying
reasoning structure that \CPP{} assumes; micro-level trace analysis is
needed.

% ================================================================
\section{Conclusion}
\label{sec:conclusion}

We introduced \CPP{} and \CIDI{} to measure how deeply LLM agents
collaborate, and showed that epistemic role framing (active-learner
stance) decisively shifts agents from expert-answering to collaborative
sense-making ($d = 1.30$ on \CDI; COUPLING more than doubles). Partner-awareness
alone does not produce this shift ($d = 0.16$, not significant after
Bonferroni correction), and C7 numerically outperforms C2 in all 30
subject-difficulty cells (descriptive; per-cell samples are small).
\AEC{} reveals that \CIDI{} creates structural interdependence ($85\%$
both-solo-zero) while only $25\%$ becomes activated epistemic necessity
under C7---a gap invisible to accuracy-based evaluation.

For the E\&D community: \CDI{} and the structural/activated distinction add diagnostic value beyond accuracy (which conflates COUPLING with TRIVIAL and misses PROD\_FAIL). Active-learner framing is an effective design intervention in the \CIDI-amenable regime, and the ablation conditions confirm its effect is synergistic with the \CIDI{} split---neither component alone suffices.

% ================================================================
\section*{Artifact Availability}
\label{sec:artifacts}

An anonymized release is available at \url{https://osf.io/2dytn/?view_only=bfda3073aca14cd2a52447994d4b8a56}. Full artifact documentation is in
Appendix~\ref{app:artifacts}. Annotations, scripts, and derived data
are released under CC BY 4.0.

% ================================================================
\section*{Acknowledgements}
This work was supported in part by [Funding agency redacted for
anonymous review]. We thank the research assistants who provided
annotation for the \CPP{} rubric. Compute was provided by [HPC
resource redacted for anonymous review].

\bibliography{references}

% ================================================================
\appendix

\section{CPP Annotation Rubric}
\label{app:rubric}

Each of the 12 cells $(P, C)$ for $P \in \{A, B, C, D\}$ and
$C \in \{1, 2, 3\}$ is scored on a 0--3 scale. We provide here the
anchoring descriptors for selected high-diagnostic cells.

\paragraph{Cell A1 (Exploring \& Understanding / Shared Understanding).}
\textbf{0}: No evidence that agents jointly discussed the problem;
one agent simply began solving. \textbf{1}: Agents acknowledged the
problem exists and restated it, but no reconciliation of perspectives.
\textbf{2}: Agents explicitly compared what each knows and flagged
gaps; at least one question about the other's packet was asked and
answered. \textbf{3}: Full joint problem model constructed; both
agents articulate a shared interpretation that differs from what
either could have produced alone, and this model visibly shapes the
subsequent solution strategy.

\paragraph{Cell D3 (Monitoring \& Reflecting / Team Organisation).}
\textbf{0}: No meta-discussion of the collaborative process or
strategy; agents proceed to output without review. \textbf{1}: One
agent flags a potential error but the flag is ignored by the partner.
\textbf{2}: Agents engage in a substantive exchange about whether
their approach is working; at least one revision of strategy is
proposed. \textbf{3}: Agents explicitly evaluate the quality of their
collaboration itself (e.g., ``I think we should have started from your
constraint, not mine---let's restart that step'') and implement a
correction.

\paragraph{Full rubric.} The complete 12-cell rubric with all anchor
descriptors, boundary case examples, and annotator guidelines is
available in the supplementary material.

\section{CIDI Generation Prompt}
\label{app:cidi_prompt}

The \CIDI{} split is generated using the following structured prompt
template with GPT-4.1 at temperature 0. The prompt instructs the
generator to:
\begin{enumerate}
  \item Identify the core mathematical objects and relations in the
  problem.
  \item Partition those objects into two groups such that group~A
  provides the structural constraints and group~B provides the
  quantitative anchors (or a domain-appropriate analogue).
  \item Verify that a solver with only group~A information cannot
  determine the numerical answer even with unlimited mathematical
  knowledge, and vice versa.
  \item Express each packet in natural language, avoiding explicit
  statements of what is absent from the other packet.
\end{enumerate}
The generated split is validated by running a single GPT-4.1 agent
on each packet alone; if the agent reaches the correct answer, the
split is regenerated with tighter constraints. Typical regeneration
rate was $12\%$ of problems.

\section{Pilot Study: Condition Selection}
\label{app:pilot}

The scale study conditions (C2, C6, C7) were selected on the basis of
a pilot study involving 7 conditions (C1--C7) run on 4 anchor
problems across 3 replications ($= 84$ conversations). C1 (no
jigsaw, no framing) was eliminated because \CDI{} was uniformly near
zero. C3 (constitutional split, negative instruction) and C5 (C3 +
peer-aware) were eliminated because agents frequently bypassed the
information restriction, invalidating the independence manipulation.
C4 (CIDI split + joint accountability framing) was eliminated due to a
design conflict: the joint accountability instruction asked agents to
share responsibility for a combined answer, which undermined the
reasoning-output dependency that the CIDI split requires. The pilot
established an ordering C7 $>$ C4 $>$ C2 $>$ C6 $>$ C3 $>$ C5 $>$
C1 for \CDI{}, motivating the three-way comparison in the scale study.
The unexpected result that C6 fell below C2 in the pilot (in the full
study, C6 scores marginally above C2, $d=0.16$, but does not survive
Bonferroni correction) first alerted us to the practically null effect
of peer-awareness.

\section{Subject-Level Generalisability Results}
\label{app:heatmap}

Table~\ref{tab:heatmap} reports mean \CDI{} for each of the 30
subject $\times$ difficulty cells under conditions C2 and C7. C7
numerically outperforms C2 in all 30 cells. Cell $n$ values range from
1 to 8 (mean 4.7), which precludes formal per-cell significance testing;
$d$ values for cells with $n \geq 3$ range from $0.63$ to $2.77$.
Effects tend to be larger at higher difficulty levels across all subjects.

\begin{table}[h]
  \caption{Cell-level \CDI{} means and Cohen's $d$ (C7 vs.\ C2) for
  all 30 subject $\times$ difficulty combinations. $d$ is omitted for
  $n < 3$ (insufficient data). Subject labels follow MATH benchmark
  categories \citep{hendrycks2021math}.}
  \label{tab:heatmap}
  \centering
  \small
  \begin{tabular}{llrrrr}
    \toprule
    \textbf{Subject} & \textbf{Level} & \textbf{$n$} &
    \textbf{CDI (C2)} & \textbf{CDI (C7)} & \textbf{$d$} \\
    \midrule
    Algebra         & 1 &  4 & 0.342 & 0.667 & 2.09 \\
    Algebra         & 2 &  3 & 0.333 & 0.711 & 2.51 \\
    Algebra         & 3 &  2 & 0.333 & 0.517 & --- \\
    Algebra         & 4 &  7 & 0.386 & 0.557 & 1.38 \\
    Algebra         & 5 &  6 & 0.378 & 0.647 & 2.06 \\
    \midrule
    Geometry        & 1 &  5 & 0.360 & 0.483 & 1.25 \\
    Geometry        & 2 &  4 & 0.325 & 0.568 & 1.60 \\
    Geometry        & 3 &  6 & 0.305 & 0.536 & 1.19 \\
    Geometry        & 4 &  6 & 0.268 & 0.644 & 1.53 \\
    Geometry        & 5 &  6 & 0.306 & 0.619 & 2.29 \\
    \midrule
    Number Theory   & 1 &  1 & 0.367 & 0.833 & --- \\
    Number Theory   & 2 &  5 & 0.326 & 0.546 & 0.73 \\
    Number Theory   & 3 &  6 & 0.290 & 0.496 & 2.77 \\
    Number Theory   & 4 &  5 & 0.364 & 0.659 & 0.90 \\
    Number Theory   & 5 &  8 & 0.409 & 0.669 & 0.91 \\
    \midrule
    Prealgebra      & 1 &  3 & 0.306 & 0.428 & 1.81 \\
    Prealgebra      & 2 &  6 & 0.403 & 0.581 & 1.14 \\
    Prealgebra      & 3 &  7 & 0.399 & 0.676 & 2.18 \\
    Prealgebra      & 4 &  4 & 0.336 & 0.558 & 1.30 \\
    Prealgebra      & 5 &  4 & 0.383 & 0.606 & 0.91 \\
    \midrule
    Precalculus     & 1 &  2 & 0.462 & 0.517 & --- \\
    Precalculus     & 2 &  3 & 0.342 & 0.606 & 1.86 \\
    Precalculus     & 3 &  6 & 0.325 & 0.508 & 0.78 \\
    Precalculus     & 4 &  7 & 0.431 & 0.717 & 1.73 \\
    Precalculus     & 5 &  5 & 0.424 & 0.737 & 1.14 \\
    \midrule
    Probability     & 1 &  1 & 0.567 & 0.817 & --- \\
    Probability     & 2 &  4 & 0.439 & 0.587 & 0.63 \\
    Probability     & 3 &  2 & 0.556 & 0.867 & --- \\
    Probability     & 4 &  5 & 0.271 & 0.609 & 1.30 \\
    Probability     & 5 &  7 & 0.375 & 0.685 & 1.38 \\
    \bottomrule
  \end{tabular}
\end{table}

% ================================================================
\section{Robustness and Validity Checks — Details}
\label{app:robustness}

\paragraph{W2 — C6 rules out partner-awareness, not ignorance.}
C6 establishes explicit partner-awareness without any learner-role
component. If simulated ignorance alone explained C7's gains, C6
should outperform C2 (since C6 agents equally lack full-solution access).
The negligible C6 vs.\ C2 effect ($d=0.16$, $\Delta_{\CDI}=+0.028$,
$p=0.026$, not surviving Bonferroni correction) rules out simple partner-awareness
as a sufficient mechanism. C6 does \emph{not} fully isolate the active-learner
stance from expert-certainty withholding; an ignorance-only ablation
(``reason cautiously; you may not have the full answer'' without
collaboration directives) remains a key future experiment.

\paragraph{W3 — CDI and correctness are orthogonal (by design).}
Across 1,967 conversations, $P(\text{correct} \mid \CDI \geq 0.5)
= 0.321$ vs.\ $P(\text{correct} \mid \CDI < 0.5) = 0.324$
($\chi^2 = 0.01$, $p = 0.94$). \CDI{} measures process quality, not
outcome quality. The C7 COUPLING advantage ($24\%$ vs.\ $11\%$ C2)
holds at identical mean conversation length (19.9 turns across all
conditions): C7 achieves $0.031$ \CDI{} per turn vs.\ $0.018$ for C2,
ruling out a length confound.

\paragraph{W4 — Convergent validity with DAG-weighted CDI.}
A topology-weighted variant $\CDI_w$ (phase depth $\times$ competency
depth, normalized) yields $\CDI_w = 0.369, 0.397, 0.601$ for C2, C6,
C7 (vs.\ uniform $0.362, 0.390, 0.613$). Spearman $\rho > 0.99$ across
all 1,967 conversations confirms that condition rankings are robust to
the specific weighting scheme.

\paragraph{W5 — Power and confidence intervals.}
Bootstrap $95\%$ CI for $\Delta_\CDI$(C7$-$C2) $= [0.219, 0.284]$,
entirely above zero. $n=140$ problems provides $>99.9\%$ power at
$d=1.30$, $\alpha=0.05$. Within-problem replication variance
($\sigma_{\text{within}} \approx 0.23$) implies single-problem \CDI{}
estimates carry substantial uncertainty ($95\%$ CI $\approx \pm 0.26$).

% ================================================================
\section{Artifact Documentation}
\label{app:artifacts}

The anonymized review release contains the following artifacts:
\begin{enumerate}
  \item \textbf{CPP rubric}: complete 12-cell annotation rubric (anchor
  descriptors, boundary cases, annotator guidelines).
  \item \textbf{CIDI generation prompt}: full structured prompt template
  and validation logic for GPT-4.1--based split generation.
  \item \textbf{Problem corpus}: 140 \CIDI{} split triples (shared
  context / Packet~A / Packet~B) identified by MATH benchmark ID.
  Original problem text is not redistributed per the MATH license;
  MATH IDs and derived splits are provided.
  \item \textbf{Conversation transcripts}: 1,967 turn-by-turn logs
  (C2, C6, C7) plus 832 ablation logs (C\textsubscript{FULL},
  C\textsubscript{EXP}).
  \item \textbf{CPP annotations}: per-cell annotator scores, raw
  disagreement records, and consensus resolutions for the 36 reliability
  conversations.
  \item \textbf{Scoring scripts}: CPP metric computation (CDI, CQI,
  PhAQ, ATC, CY) and quadrant assignment.
  \item \textbf{AEC scripts}: solo-run correctness evaluation and
  Shapley-value computation.
  \item \textbf{Benchmark card}: corpus construction, selection
  conditionality, intended use, known limitations, and dataset metadata.
\end{enumerate}
Annotations, scripts, and derived data: CC BY 4.0.
Conversation logs: terms of the applicable API provider.
A Croissant metadata file is included for E\&D dataset submission
requirements. A reproduction README documents the full pipeline from
problem selection through scoring.

% ================================================================
\section*{NeurIPS Paper Checklist}

\begin{enumerate}

\item \textbf{Claims.} Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
\answerYes{} All quantitative claims are directly supported by reported results. CDI gains, effect sizes, and quadrant percentages are computed from 1,967 conversations. Ablation claims are supported by 832 additional conversations (Section~\ref{sec:ablations}). All results are described as conditional on the CIDI-amenable corpus (§4.1, L1 in Limitations).

\item \textbf{Limitations.} Does the paper discuss the limitations of the work?
\answerYes{} Six limitations are explicitly labeled (L1--L6) in the Limitations paragraph. L1 (corpus conditionality) is the most critical and is noted in the abstract, §4.1, and §6. L2 identifies remaining ablation gaps. L3 clarifies AEC scope.

\item \textbf{Theory Assumptions and Proofs.} Does the paper make theoretical claims?
\answerNA{} This is an empirical benchmark and evaluation paper. No theoretical proofs are made.

\item \textbf{Experimental Result Reproducibility.} Does the paper fully describe the conditions necessary for an experimenter to reproduce the main results?
\answerYes{} Full experimental conditions are described in §4. Model, temperature, prompt structure, corpus, and annotation protocol are documented. Scoring scripts, prompts, and conversation logs are provided in the artifact release (Appendix~\ref{app:artifacts}).

\item \textbf{Open access to data and code.} Does the paper provide open access to the data and code used?
\answerYes{} An anonymized review release provides CPP rubric, CIDI prompt, problem corpus (MATH IDs + derived splits), transcripts, annotations, scoring scripts, AEC scripts, and benchmark card. Full documentation in Appendix~\ref{app:artifacts}.

\item \textbf{Experimental Setting/Details.} Does the paper specify all the experimental settings/hyperparameters important to reproduce the main results?
\answerYes{} Model (GPT-4.1), temperature ($T = 0$ for solo AEC runs; $T = 0.7$ with different random seeds for collaborative conversations), maximum turns (20), replication count (3--7 per problem-condition), and condition prompts are all specified or provided in the artifact.

\item \textbf{Experiment Statistical Significance.} Does the paper report error bars suitably and correctly?
\answerYes{} SE reported for all CDI means. Bootstrap CIs reported for the main C7--C2 comparison. Bonferroni correction applied for multiple condition comparisons. Effect sizes (Cohen's $d$) reported for all pairwise contrasts. Power analysis in Appendix~\ref{app:robustness}.

\item \textbf{Experiments Compute Resources.} Does the paper discuss the computational resources used to run the experiments?
\answerYes{} All conversations were generated via the OpenAI API. Compute was provided by [HPC resource redacted for review]. Approximate total cost is in the range \$50--\$200 USD for all conditions including ablations.

\item \textbf{Code Of Ethics.} Have you read the NeurIPS Code of Ethics?
\answerYes{}

\item \textbf{Broader Impacts.} Does the paper discuss both potential positive societal impacts and negative societal impacts of the work?
\answerYes{} The work improves evaluation of AI collaboration quality, with positive implications for education and human-AI teaming. Risks include overfitting evaluation design to the CPP rubric; we discuss this under L6 (collaborative register vs.\ reasoning). No sensitive data or dual-use concerns apply.

\item \textbf{Licenses for existing assets.} Does the paper cite the creators of assets it uses and include appropriate license information?
\answerYes{} MATH benchmark \citep{hendrycks2021math} is used under its original license; original problem text is not redistributed. PISA 2015 CPS framework \citep{pisa2015cps} is cited. All derived artifacts are CC BY 4.0.

\item \textbf{New Assets.} Are new assets introduced in the paper documented appropriately?
\answerYes{} The CPP rubric, CIDI prompt, conversation corpus, annotation data, and scripts are documented in Appendix~\ref{app:artifacts} and the benchmark card.

\item \textbf{Crowdsourcing and Research with Human Subjects.} Does the paper involve crowdsourcing or research with human subjects?
\answerYes{} Graduate-student annotators provided inter-rater reliability annotations (§4.3). Annotation was conducted under standard research practice; no personal data were collected from participants.

\item \textbf{Institutional Review Board (IRB) Approvals.} Does the paper describe potential risks incurred by study participants?
\answerYes{} Annotation involved graduate research assistants scoring model-generated mathematical text; no personal data were collected. The activity was treated as minimal-risk annotation under applicable institutional practice, and participants were compensated for their time. No formal IRB review was required under these conditions.

\end{enumerate}

\end{document}
